---
title: "Basics of data import and cleaning in pathviewR"
author: "Vikram B. Baliga"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics of data import and cleaning in pathviewR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{tidyverse}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Data import via `pathviewR`

Our `read_` functions presently support `.csv` files exported from 
[Optitrack's Motive](https://optitrack.com/products/motive/) software 
as well as `.mat` files exported from 
[Flydra](https://github.com/strawlab/flydra).

Should you have data from other sources, please see the function `as_viewr()`
and the ACCOMPANYING VIGNETTE (LINK!).

We'll start by loading `pathviewR` and `tidyverse`.

```{r package_loading, message=FALSE, warning=FALSE}
## If you do not already have pathviewR installed:
# install.packages("devtools")
# devtools::install_github("vbaliga/pathviewR")

library(pathviewR)
library(tidyverse)
```

### Motive CSV files

`.csv` files exported from Motive can be imported via `read_motive_csv()`

```{r import_motive}
## Import the Motive example data included in 
## the package

motive_data <-
  read_motive_csv(
    system.file("extdata", "pathviewR_motive_example_data.csv",
                package = 'pathviewR')
  )

## This produces a tibble 
motive_data
``` 

Metadata are stored as attributes. We won't go through all of these, but here
are a couple important ones.

``` {r metadata}
## E.g. to see the header of the original file:
attr(motive_data, "header")

## Names of all marked objects:
attr(motive_data, "subject_names_simple")

## Types of data included
attr(motive_data, "data_types_simple")

## Frame rate
attr(motive_data, "frame_rate")
```

### Flydra Matlab files

`.mat` files exported from Flydra can be imported via `read_flydra_mat()`.

Note that a subject name is required. Only one name can be added and it will
be used throughout the resultant `tibble`.

```{r import_flydra}
## Import the Flydra example data included in 
## the package
flydra_data <-
  read_flydra_mat(
    system.file("extdata", 
                "pathviewR_flydra_example_data.mat",
                package = 'pathviewR'),
    subject_name = "birdie_wooster"
  )

## Similarly, this produces a tibble with important 
## metadata as attributes
flydra_data
```

## Data cleaning
Data exported via either Motive or Flydra are not typically "tidy". Functions in
`pathviewR` ultimately rely on having tidy data sets that are easily 
interpreted.

Several functions to clean and wrangle data are available, and we have a
suggested pipeline for how these steps should be handled. The rest of this
vignette will cover these steps.

All of the steps in the suggested pipeline are also covered by two all-in-one
functions: `clean_viewr()` and `import_and_clean_viewr()`. See the section at
the very end of this vignette for details.

And speaking of pipes, all functions in `pathviewR` are pipe-friendly. We will
detail each step separately, but each of the subsequent steps may be piped, e.g.
`data %>% relabel_viewr_axes() %>% gather_tunnel_data()` etc etc

### Relabeling axes, gathering data columns, and trimming outliers
Axis labels (x, y, z) may be applied in arbitrary ways by software. A user might
intuitively think the z axis represents height, but the original software may
label it as the y axis instead.

`relabel_viewr_axes` provides a means to relabel axes with  "tunnel_length",
"tunnel_width", and "tunnel_height". **These axis labels will be expected by
subsequent functions, so skipping this step is ill-advised.**

```{r relabel_axes}
motive_relabeled <-
  motive_data %>%
  relabel_viewr_axes(
    tunnel_length = "_z",
    tunnel_width = "_x",
    tunnel_height = "_y",
    real = "_w"
  )

names(motive_relabeled)
```

Akin to the behavior of `dplyr::gather()`, `gather_tunnel_data()` will take all
data from a given session and organize it so that all data of a given type are
within one column, i.e. all position lengths are in `position_length`, as
opposed to separate length columns for each rigid body. **These column names
will be expected by subsequent functions, so skipping this step is also
ill-advised.**

Use `trim_tunnel_outliers()` to remove artifacts and other outlier data. This
step is entirely optional, and should only be used when the user is confident
that data outside certain ranges are artifacts or other bugs. Data outside these
ranges are then filtered out. Best to plot data beforehand and check!!

```{r gather_and_trim}
## First gather and show the new column names
motive_gathered <-
  motive_relabeled %>%
  gather_tunnel_data()

names(motive_gathered)

## Now trim, using ranges we know to safely include data
## and exclude artifacts. Anything outside these ranges 
## will be removed.
motive_trimmed <-
  motive_gathered %>%
  trim_tunnel_outliers(
    lengths_min = 0,
    lengths_max = 3,
    widths_min = -0.4,
    widths_max = 0.8,
    heights_min = -0.2,
    heights_max = 0.5
  )
```

### Standardization of tunnel position and coordinates
The coordinate system of the tunnel itself may require adjustment or
standardization. For example, data collected across different days may show
slight differences in coordinate systems if calibration equipment was not used
in identical ways. Moreover, the user may want to redefine how the coordinate
system itself is defined (i.e. change the location of `(0, 0, 0)` to another
place within the tunnel.

Note that having `(0, 0, 0)` set to the center of the tunnel is required for 
all subsequent `pathviewR` functions to work.

`pathviewR` offers three main choices for such standardization:  

- `redefine_tunnel_center()`: Sets the location of 0 on any or all axes to a new
location. See the Help page for this function to see the four different methods
by which a user can specify this. No rotation of the tunnel is performed.  

- `standardize_tunnel()`: Use specified landmarks (`subjects` within the `viewr`
object) to rotate and translate the location of a tunnel, setting `(0, 0, 0)` to
the center of the tunnel (centering).

- `rotate_tunnel`:  Rotate and center a tunnel based on user-defined coordinates
(i.e. similar to `standardize_tunnel()` but for cases where specified landmarks
are not in the data).  

Two quick examples will follow, using our motive and Flydra data:

```{r rotate_example}
## Rotate and center the motive data set:
motive_rotated <-
  motive_trimmed %>% 
  rotate_tunnel(
    perch1_len_min = -0.06,
    perch1_len_max = 0.06,
    perch2_len_min = 2.48,
    perch2_len_max = 2.6,
    perch1_wid_min = 0.09,
    perch1_wid_max = 0.31,
    perch2_wid_min = 0.13,
    perch2_wid_max = 0.35
  )
```

In the above, virtual perches are defined by the user using the arguments shown.
The center of each perch is then found and then the locations of the two perch
centers are then used to 1) set  `(0, 0, 0)` to the point that is equidistant
from the perches (i.e. the middle of the tunnel) and 2) rotate the tunnel about
the height axis so that both perch width coordinates are at 0. This may be
easier to understand through plotting:

```{r rotate_example_plots}
## Quick (base-R) plot of the original data
plot(motive_trimmed$position_length,
     motive_trimmed$position_width,
     asp = 1)

## Quick (base-R) plot of the rotated data
plot(motive_rotated$position_length,
     motive_rotated$position_width,
     asp = 1)
```

Differences due to rotation may be extremely subtle, but the redefining of 
`(0, 0, 0)` to the middle of the tunnel should be clear from contrasting the
axes of the plots.

Flydra data typically do not need to be rotated, so we will instead use
`redfine_tunnel_center()` to adjust the location of `(0, 0, 0)`:

```{r redefine_tunnel_example}
## Re-center the Flydra data set:
flydra_centered <-
  flydra_data %>%
  redefine_tunnel_center(length_method = "middle",
                         height_method = "user-defined",
                         height_zero = 1.44)
```

Here, we are using `length_method = "middle"` to use the middle of the range of
"length" data to set length = 0 (a translation), making no change to the width
axis, and then specifying that height = 0 should be equal to the value `1.44`
from the original data (another translation). Again, plotting may help; note
that this time, we'll plot length x height (instead of width):

```{r redefine_example_plots}
## Quick (base-R) plot of the original data
plot(flydra_data$position_length,
     flydra_data$position_height,
     asp = 1)

## Quick (base-R) plot of the redefined data
plot(flydra_centered$position_length,
     flydra_centered$position_height,
     asp = 1)
```

### Selecting a region of interest
This required step has benefits that are twofold: 1) treatment effects on animal
movement may only manifest over certain regions of the tunnel, and 2) focusing
on a subset of the data makes it easier to define explicit trajectories and run
computations faster.

The region of interest is defined via the function `select_x_percent()`. Once
tunnel coordinates have been standardized (via one of the function in the
previous section), `select_x_percent()` then selects the middle `x` percent
(along the length axis) of the tunnel as the region of interest. For example,
selecting 50 percent would start from the center of the tunnel and move 25% of
the tunnel along positive length and 25% along negative length values to then
select the middle 50% of the tunnel.

Quick examples:
```{r select_x_examples}
## Motive data: select the middle 50% of the tunnel as the region of interest
motive_selected <-
  motive_rotated %>% select_x_percent(50)

## Quick plot:
plot(motive_selected$position_length,
     motive_selected$position_width,
     asp = 1)

## Flydra data:
flydra_selected <-
  flydra_centered %>% select_x_percent(50)

## Quick plot:
plot(flydra_selected$position_length,
     flydra_selected$position_width,
     asp = 1)
```

### Isolating each trajectory
The `pathviewR` standard for defining a trajectory is: continuous movement from
one side of the tunnel to the other over the span of the region of interest.
Note that this definition does not strictly require linear movement from one end
to the other; an animal could make several loops inside the region of interest
within a given trajectory.  

Isolating trajectories is handled via the `separate_trajectories()` function in
`pathviewR`. Note that a region of interest must be selected beforehand via
`select_x_percent()`.  

Because cameras may occasionally drop frames, we allow the user to permit some
relaxation of how stringent the "continuous movement" criterion is. This is
handled via the `max_frame_gap` argument within `separate_trajectories()`. For
more details, please see VIGNETTE XXX (LINK HERE).  

In our Motive example, we'll use the automated feature built into the function
to guesstimate the best `max_frame_gap` allowed. When frame gaps larger than
`max_frame_gap` are encountered, the function will force the defining of a new
trajectory. But if frame gaps smaller than `max_frame_gap` are encountered,
keeping observations within the same trajectory is permitted.

In the Flydra example, we'll simply set `max_frame_gap` to `1` so that no frame
gaps are allowed (movement must be continuous with no dropped frames).

```{r separate_examples}
motive_labeled <-
  motive_selected %>% 
  separate_trajectories(max_frame_gap = "autodetect")

flydra_labeled <-
  flydra_selected %>% 
  separate_trajectories(max_frame_gap = 1)
```

### Retain only complete trajectories
Now that trajectories have been isolated and labeled, the final cleaning step is
to retain only the trajectories that completely span from one end of the region
of interest to the other.

This final step is handled via `pathviewR`'s `get_full_trajectories()`.

There is a built-in "fuzziness" feature: because trajectories may not have
observations exactly at the beginning or the end of the region of interest, it
may be necessary to allow trajectories to be slightly shorter than the range of
the selected region of interest. The `span` parameter of this function handles
this. By supplying a numeric proportion from `0` to `1`, a user may allow
trajectories to span that proportion of the selected region. For example,
setting `span = 0.95` will keep all trajectories that span 95% of the length of
the selected region of interest. Setting `span = 1` (not recommended) will
strictly keep trajectories that start and end at the **exact** cut-offs of the
selected region of interest.

For these reasons, `span`s of `0.99` to `0.95` are generally recommended. The
best choice ultimately depends on your capture frame rate as well as your own
judgment. Should you desire to set it lower (which you can), you may instead
consider using a smaller region of interest (i.e. set the `desired_percent`
parameter in `select_x_percent()` to be lower).

```{r get_full_examples}
## Motive
motive_full <-
  motive_labeled %>% 
  get_full_trajectories(span = 0.95)

plot(motive_full$position_length,
     motive_full$position_width,
     asp = 1, col = as.factor(motive_full$file_sub_traj))

## Flydra
flydra_full <-
  flydra_labeled %>% 
  get_full_trajectories(span = 0.95)

plot(flydra_full$position_length,
     flydra_full$position_width,
     asp = 1, col = as.factor(flydra_full$file_sub_traj))
```

### All-in-one cleaning functions
All of the above steps can also be done by using `pathviewR`'s designated 
all-in-one functions. `import_and_clean_viewr()` imports raw data and allows
the user to run through all of the cleaning steps previously covered in this
vignette. `clean_viewr()` does the same on any object already imported into the
R environment (i.e. it skips data import).

For both of these functions, all of the cleaning steps are set to `TRUE` by 
default, but may be turned off by using `FALSE`. Argument names correspond to 
standalone functions in `pathviewR`, and if the user wants to use non-default values for corresponding arguments, they should also be supplied for any steps 
that are set to `TRUE`. 

For example, if the user keeps `select_x_percent = TRUE` as an argument in
`clean_viewr()`, the `select_x_percent()` function is run internally. This means
that should the user desire to select a region of interest that does not match
the default value of 33 percent, an additional argument should be supplied to
`clean_viewr()` as if it were being supplied to `select_x_percent()` itself, 
e.g.: `desired_percent = 50`.

All additional arguments should be written out fully and explicitly.

Here's an example using what we previously covered:

```{r all-in-one}
motive_allinone <-
  motive_data %>%
  clean_viewr(
    relabel_viewr_axes = TRUE,
    gather_tunnel_data = TRUE,
    trim_tunnel_outliers = TRUE,
    standardization_option = "rotate_tunnel",
    select_x_percent = TRUE,
    desired_percent = 50,
    rename_viewr_characters = FALSE,
    separate_trajectories = TRUE,
    max_frame_gap = "autodetect",
    get_full_trajectories = TRUE,
    span = 0.95
  )

## Quick plot
plot(motive_allinone$position_length,
     motive_allinone$position_width,
     asp = 1, col = as.factor(motive_allinone$file_sub_traj))
```


That's all!

🐢
