---
title: "Basics of data import and cleaning in pathviewr"
author: "Vikram B. Baliga"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics of data import and cleaning in pathviewr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{magrittr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Raw movement data, including those from motion capture systems, may have a
variety of issues. These raw data often contain noise or artifacts from the
recording session, which may not be easily removed via the recording software
itself. Data may not be organized as “tidy” key-value pairs (making plotting
more difficult), the axes and overall orientation of the environment may not
conform to a standard, and individual movement trajectories may be ill-defined.

`pathviewr` provides functions in R to deal with such problems (i.e.
"cleaning"). This vignette will cover the basics of how to import raw data and
how to clean data to prepare it for visualization and/or statistical analyses.


## What do movement data sets look like?

At minimum, movement data provide information on a subject or object's position
over time. These data are typically supplied in three dimensions (e.g. x, y, z),
with position in each dimension sampled at a particular rate (e.g. 100 Hz).
Different recording software may provide additional features, such as the
ability to track multiple subjects simultaneously, information on subjects'
rotation, tracking of "rigid body" elements, or even the ability to apply Kalman
filters.

A central goal of `pathviewr` is to take data from different sources (so far:
Motive and Flydra), re-organize them into a common format that can be wrangled
in R, clean them up a bit, and get them ready for visualization and/or
statistical analyses. We'll first cover what's included in Motive and in Flydra
data and how `pathviewr` handles these. Should you have data from another
source, our `as_viewr()` function will allow you to bring it into the
`pathviewr` framework.

## Data import via `pathviewr`

Data can be imported via one of three functions:  

- `read_motive_csv()` imports data from `.csv` files that have been exported
from Optitrack's [Motive](https://optitrack.com/software/motive/) software

- `read_flydra_mat()` imports data from  `.mat` files that have been exported
from [Flydra](https://github.com/strawlab/flydra)

- `as_viewr()` can be used to handle data from other sources  

We will showcase examples from each of these methods in this section.

Please feel free to reach out to the `pathviewr` authors via 
[our Github Issues page](https://github.com/vbaliga/pathviewr/issues/) 
should you have trouble with any of our data import options. We are happy to
work with you to design custom `read_` functions for file types we have not
encountered ourselves. 

We'll start by loading `pathviewr` and a few of the packages in the `tidyverse`.

```{r package_loading, message=FALSE, warning=FALSE}
## If you do not already have pathviewr installed:
# install.packages("devtools")
# devtools::install_github("vbaliga/pathviewr")

library(pathviewr)
library(ggplot2)
library(magrittr)
```

### Motive CSV files

`.csv` files exported from Motive can be imported via `read_motive_csv()`

```{r import_motive}
## Import the Motive example data included in 
## the package

motive_data <-
  read_motive_csv(
    system.file("extdata", "pathviewr_motive_example_data.csv",
                package = 'pathviewr')
  )

## This produces a tibble 
motive_data
``` 

A key thing to note is that these data, as stored in Motive CSVs, are not
"tidy". Each frame occupies one row, but what that also means is that the
rotation and position values for the various subjects take up 24 columns! This
format not only makes plotting data more difficult in base R, `ggplot2`, and
`rgl`, but also makes other aspects of data wrangling more difficult. In a later
step, we will 'gather' these data into key-value pairs so that e.g. all
length-wise position values are in one column, all width-wise are in
another...etc.

Metadata are stored as attributes. We won't go through all of these, but here
are a couple important ones.

``` {r metadata}
## E.g. to see the header of the original file:
attr(motive_data, "header")

## Names of all marked objects:
attr(motive_data, "subject_names_simple")

## Types of data included
attr(motive_data, "data_types_simple")

## Frame rate
attr(motive_data, "frame_rate")
```

Storing such metatdata in the attributes is a key feature of `pathviewr`. These
metadata may not be as immediately as important as the time series of position
or rotation, but they can provide important experimental information such as the
date & time of capture and the units of the position data (here, meters).

### Flydra Matlab files

`.mat` files exported from Flydra can be imported via `read_flydra_mat()`.

Note that you must supply a `subject_name` for Flydra data, as subject names 
are not embedded in the `.mat` files. Only one name can be added and it will
be used throughout the resultant `tibble`.

```{r import_flydra}
## Import the Flydra example data included in 
## the package
flydra_data <-
  read_flydra_mat(
    system.file("extdata", 
                "pathviewr_flydra_example_data.mat",
                package = 'pathviewr'),
    subject_name = "birdie_wooster"
  )

## Similarly, this produces a tibble with important 
## metadata as attributes
flydra_data

attr(flydra_data, "frame_rate")
```

Note that unlike the example Motive data, the Flydra data are already organized
into key-value pairs. Because rotation is not captured by Flydra, such data are
also not included.

### Data from other sources

Data from another format can be converted to a `viewr` object via
`pathviewr::as_viewr()`. Although this function does not handle data import
per se, it allows data that you may already have imported into R as a `tibble`
or `data.frame` to then be reformatted for use with `pathviewr` functions.

We'll run through a quick example with simulated data:

```{r as_viewr}
## Create a dummy data frame with simulated (nonsense) data
df <-
  data.frame(
    frame = seq(1, 100, by = 1),
    time_sec = seq(0, by = 0.01, length.out = 100),
    subject = "birdie_sanders",
    z = rnorm(100),
    x = rnorm(100),
    y = rnorm(100)
  )

## Use as_viewr() to convert it into a viewr object
test <-
  as_viewr(
    df,
    frame_rate = 100,
    frame_col = 1,
    time_col = 2,
    subject_col = 3,
    position_length_col = 5,
    position_width_col = 6,
    position_height_col = 4
  )

## Some metadata are stored as attributes
attr(test, "frame_rate")
```

We also welcome you to request custom data import functions, especially if
`as_viewr()` does not fit your needs. We are interested in expanding our data
import functions to accommodate additional file types. Please feel free to 
[file a request for additional import functions](https://github.com/vbaliga/pathviewr/issues/new/choose) 
via our Github Issues page.


## Data cleaning
As noted above, raw data often suffer the following:  
- contain noise or artifacts from the recording session  
- not organized as “tidy” key-value pairs  
- axes and overall orientation of the environment may not conform to a standard  
- individual movement trajectories may be ill-defined

Several functions to clean and wrangle data are available, and we have a
suggested pipeline for how these steps should be handled. The rest of this
vignette will cover these steps.

All of the steps in the suggested pipeline are also covered by two all-in-one
functions: `clean_viewr()` and `import_and_clean_viewr()`. See the section at
the very end of this vignette for details.

And speaking of pipes, all functions in `pathviewr` are pipe-friendly. We will
detail each step separately, but each of the subsequent steps may be piped, e.g.
`data %>% relabel_viewr_axes() %>% gather_tunnel_data()` etc etc

### Relabeling axes, gathering data columns, and trimming outliers
Axis labels (x, y, z) may be applied in arbitrary ways by software. A user might
intuitively think the z axis represents height, but the original software may
label it as the y axis instead.

`relabel_viewr_axes` provides a means to relabel axes with "tunnel_length",
"tunnel_width", and "tunnel_height". **These axis labels will be expected by
subsequent functions, so skipping this step is ill-advised.**

Typically, axes from Motive data will need to be relabled, but axes in data
imported from Flydra will not.

```{r relabel_axes}
motive_relabeled <-
  motive_data %>%
  relabel_viewr_axes(
    tunnel_length = "_z",
    tunnel_width = "_x",
    tunnel_height = "_y",
    real = "_w"
  )

names(motive_relabeled)
```

Akin to the behavior of `dplyr::gather()`, `gather_tunnel_data()` will take all
data from a given session and organize it so that all data of a given type are
within one column, i.e. all position lengths are in `position_length`, as
opposed to separate length columns for each rigid body. **These column names
will be expected by subsequent functions, so skipping this step is also
ill-advised if you are using data from Motive.** Should you have data from 
Flydra, this step should be skipable.

Use `trim_tunnel_outliers()` to remove extreme artifacts and other outlier data.
What this function does is create a (virtual) boundary box according to
user-specification, and any data outside that boundary are removed. For example,
if you know your arena measures 10m x 10m x 10m and your data were calibrated to
range from 0-10m in each dimension, you can be reasonably sure that extreme
values such as 45m on a given axis are bogus. This step is entirely optional,
and should only be used when the user is confident that data outside certain
ranges are artifacts or other bugs. Data outside these ranges are then filtered
out. Best to plot data beforehand and check!!

```{r gather_and_trim}
## First gather and show the new column names
motive_gathered <-
  motive_relabeled %>%
  gather_tunnel_data()

names(motive_gathered)

## Now trim, using ranges we know to safely include data
## and exclude artifacts. Anything outside these ranges 
## will be removed.
motive_trimmed <-
  motive_gathered %>%
  trim_tunnel_outliers(
    lengths_min = 0,
    lengths_max = 3,
    widths_min = -0.4,
    widths_max = 0.8,
    heights_min = -0.2,
    heights_max = 0.5
  )
```

### Standardization of tunnel position and coordinates
The coordinate system of the tunnel itself may require adjustment or
standardization. For example, data collected across different days may show
slight differences in coordinate systems if calibration equipment was not used
in identical ways. Moreover, the user may want to redefine how the coordinate
system itself is defined (i.e. change the location of `(0, 0, 0)` to another
place within the tunnel.

Note that having `(0, 0, 0)` set to the center of the region of interest
(covered in the next section of this vignette) is required for all subsequent
`pathviewr` functions to work.

`pathviewr` offers three main choices for such standardization:  

- `redefine_tunnel_center()`: Sets the location of 0 on any or all axes to a new
location. See the Help page for this function to see the four different methods
by which a user can specify this. No rotation of the tunnel is performed. This
function can be used on both Motive and Flydra data.  

- `standardize_tunnel()`: Use specified landmarks (`subjects` within the `viewr`
object) to rotate and translate the location of a tunnel, setting `(0, 0, 0)` to
the center of the tunnel (centering). For example, in an avian flight tunnel,
perches may be set up on opposite ends of the tunnel and rigid body markers may
be set to them. The positions of these perches can be used as landmarks to
standardize tunnel position. Note that this is typically not possible for Flydra
data, since Flydra data will be imported with only one `subject`.  

- `rotate_tunnel`: Rotate and center a tunnel based on user-defined coordinates
(i.e. similar to `standardize_tunnel()` but for cases where specified landmarks
are not in the data). This function can be used on both Motive and Flydra data.  

Two quick examples will follow, using our Motive and Flydra data:

```{r rotate_example}
## Rotate and center the motive data set:
motive_rotated <-
  motive_trimmed %>% 
  rotate_tunnel(
    perch1_len_min = -0.06,
    perch1_len_max = 0.06,
    perch2_len_min = 2.48,
    perch2_len_max = 2.6,
    perch1_wid_min = 0.09,
    perch1_wid_max = 0.31,
    perch2_wid_min = 0.13,
    perch2_wid_max = 0.35
  )
```

In the above, virtual perches are defined by the user using the arguments shown.
The center of each perch is then found and then the locations of the two perch
centers are then used to 1) set  `(0, 0, 0)` to the point that is equidistant
from the perches (i.e. the middle of the tunnel) and 2) rotate the tunnel about
the height axis so that both perch width coordinates are at 0. This may be
easier to understand through plotting:

```{r rotate_example_plots}
## Quick (base-R) plot of the original data
plot(motive_trimmed$position_length,
     motive_trimmed$position_width,
     asp = 1)

## Quick (base-R) plot of the rotated data
plot(motive_rotated$position_length,
     motive_rotated$position_width,
     asp = 1)
```

Differences due to rotation may be extremely subtle, but the redefining of 
`(0, 0, 0)` to the middle of the tunnel should be clear from contrasting the
axes of the plots.

Flydra data typically do not need to be rotated, so we will instead use
`redefine_tunnel_center()` to adjust the location of `(0, 0, 0)`:

```{r redefine_tunnel_example}
## Re-center the Flydra data set:
flydra_centered <-
  flydra_data %>%
  redefine_tunnel_center(length_method = "middle",
                         height_method = "user-defined",
                         height_zero = 1.44)
```

Here, we are using `length_method = "middle"` to use the middle of the range of
"length" data to set length = 0 (a translation), making no change to the width
axis, and then specifying that height = 0 should be equal to the value `1.44`
from the original data (another translation). Again, plotting may help; note
that this time, we'll plot length x height (instead of width):

```{r redefine_example_plots}
## Quick (base-R) plot of the original data
plot(flydra_data$position_length,
     flydra_data$position_height,
     asp = 1)

## Quick (base-R) plot of the redefined data
plot(flydra_centered$position_length,
     flydra_centered$position_height,
     asp = 1)
```

### Selecting a region of interest
This required step has benefits that are twofold: 1) treatment effects on animal
movement may only manifest over certain regions of the tunnel, and 2) focusing
on a subset of the data makes it easier to define explicit trajectories and run
computations faster.

The region of interest is defined via the function `select_x_percent()`. Once
tunnel coordinates have been standardized (via one of the function in the
previous section), `select_x_percent()` then selects the middle `x` percent
(along the length axis) of the tunnel as the region of interest. For example,
selecting 50 percent would start from the center of the tunnel and move 25% of
the tunnel along positive length and 25% along negative length values to then
select the middle 50% of the tunnel.

Quick examples:
```{r select_x_examples}
## Motive data: select the middle 50% of the tunnel as the region of interest
motive_selected <-
  motive_rotated %>% select_x_percent(50)

## Quick plot:
plot(motive_selected$position_length,
     motive_selected$position_width,
     asp = 1)

## Flydra data:
flydra_selected <-
  flydra_centered %>% select_x_percent(50)

## Quick plot:
plot(flydra_selected$position_length,
     flydra_selected$position_width,
     asp = 1)
```

### Isolating each trajectory
The `pathviewr` standard for defining a trajectory is: continuous movement from
one side of the tunnel to the other over the span of the region of interest.
Note that this definition does not strictly require linear movement from one end
to the other; an animal could make several loops inside the region of interest
within a given trajectory.  

Isolating trajectories is handled via the `separate_trajectories()` function in
`pathviewr`. Note that a region of interest must be selected beforehand via
`select_x_percent()`.  

Because cameras may occasionally drop frames, we allow the user to permit some
relaxation of how stringent the "continuous movement" criterion is. This is
handled via the `max_frame_gap` argument within `separate_trajectories()`. For
more details, please see [the vignette Managing frame gaps with pathviewr](https://vbaliga.github.io/pathviewr/articles/managing-frame-gaps.html).  

In our Motive example, we'll use the automated feature built into the function
to guesstimate the best `max_frame_gap` allowed. When frame gaps larger than
`max_frame_gap` are encountered, the function will force the defining of a new
trajectory. But if frame gaps smaller than `max_frame_gap` are encountered,
keeping observations within the same trajectory is permitted.

In the Flydra example, we'll simply set `max_frame_gap` to `1` so that no frame
gaps are allowed (movement must be continuous with no dropped frames).

```{r separate_examples}
motive_labeled <-
  motive_selected %>% 
  separate_trajectories(max_frame_gap = "autodetect")

flydra_labeled <-
  flydra_selected %>% 
  separate_trajectories(max_frame_gap = 1)
```

### Retain only complete trajectories
Now that trajectories have been isolated and labeled, the final cleaning step is
to retain only the trajectories that completely span from one end of the region
of interest to the other.

This final step is handled via `pathviewr`'s `get_full_trajectories()`.

There is a built-in "fuzziness" feature: because trajectories may not have
observations exactly at the beginning or the end of the region of interest, it
may be necessary to allow trajectories to be slightly shorter than the range of
the selected region of interest. The `span` parameter of this function handles
this. By supplying a numeric proportion from `0` to `1`, a user may allow
trajectories to span that proportion of the selected region. For example,
setting `span = 0.95` will keep all trajectories that span 95% of the length of
the selected region of interest. Setting `span = 1` (not recommended) will
strictly keep trajectories that start and end at the **exact** cut-offs of the
selected region of interest.

For these reasons, `span`s of `0.99` to `0.95` are generally recommended. The
best choice ultimately depends on your capture frame rate as well as your own
judgment. Should you desire to set it lower (which you can), you may instead
consider using a smaller region of interest (i.e. set the `desired_percent`
parameter in `select_x_percent()` to be lower).

```{r get_full_examples}
## Motive
motive_full <-
  motive_labeled %>% 
  get_full_trajectories(span = 0.95)

plot(motive_full$position_length,
     motive_full$position_width,
     asp = 1, col = as.factor(motive_full$file_sub_traj))

## Flydra
flydra_full <-
  flydra_labeled %>% 
  get_full_trajectories(span = 0.95)

plot(flydra_full$position_length,
     flydra_full$position_width,
     asp = 1, col = as.factor(flydra_full$file_sub_traj))
```

### All-in-one cleaning functions
All of the above steps can also be done by using `pathviewr`'s designated 
all-in-one functions. `import_and_clean_viewr()` imports raw data and allows
the user to run through all of the cleaning steps previously covered in this
vignette. `clean_viewr()` does the same on any object already imported into the
R environment (i.e. it skips data import).

For both of these functions, all of the cleaning steps are set to `TRUE` by 
default, but may be turned off by using `FALSE`. Argument names correspond to 
standalone functions in `pathviewr`, and if the user wants to use non-default values for corresponding arguments, they should also be supplied for any steps 
that are set to `TRUE`. 

For example, if the user keeps `select_x_percent = TRUE` as an argument in
`clean_viewr()`, the `select_x_percent()` function is run internally. This means
that should the user desire to select a region of interest that does not match
the default value of 33 percent, an additional argument should be supplied to
`clean_viewr()` as if it were being supplied to `select_x_percent()` itself, 
e.g.: `desired_percent = 50`.

All additional arguments should be written out fully and explicitly.

Here's an example using what we previously covered:

```{r all-in-one}
motive_allinone <-
  motive_data %>%
  clean_viewr(
    relabel_viewr_axes = TRUE,
    gather_tunnel_data = TRUE,
    trim_tunnel_outliers = TRUE,
    standardization_option = "rotate_tunnel",
    select_x_percent = TRUE,
    desired_percent = 50,
    rename_viewr_characters = FALSE,
    separate_trajectories = TRUE,
    max_frame_gap = "autodetect",
    get_full_trajectories = TRUE,
    span = 0.95
  )

## Quick plot
plot(motive_allinone$position_length,
     motive_allinone$position_width,
     asp = 1, col = as.factor(motive_allinone$file_sub_traj))
```


That's all!

🐢
