---
title: "Managing frame gaps with pathviewR"
author: "Melissa S. Armstrong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing frame gaps with pathviewR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{tidyverse}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Defining trajectories with `pathviewR`
`pathviewR` defines trajectories as continuous movement from one side of a
region of interest to the other. Before trajectories can be defined, the region
of interest must be selected via `select_x_percent()` in addition to all the
previous steps of the data import and cleaning pipeline as described in [the Data Import and Cleaning vignette](https://vbaliga.github.io/pathviewR/articles/data-import-cleaning.html).
We'll start with loading `pathviewR` and `tidyverse` and importing our data.

```{r package_loading, message=FALSE, warning=FALSE}
## If you do not already have pathviewR installed:
# install.packages("devtools")
# devtools::install_github("vbaliga/pathviewR")

library(pathviewR)
library(tidyverse)

## Import the example Motive data included in 
## the package
motive_data <-
  read_motive_csv(
    system.file("extdata", "pathviewR_motive_example_data.csv",
                package = 'pathviewR')
  )
```

We'll quickly run through the cleanup pipeline using one of `pathviewR`'s
all-in-one cleaning functions as described in [the Data Import and Cleaning vignette](https://vbaliga.github.io/pathviewR/articles/data-import-cleaning.html).
Since trajectories are defined in the `separate_trajectories()` function, we'll
stop the all-in-one there by setting it and every step after it to `FALSE` so we
can take a closer look at how trajectories are defined. Since all steps are set
to `TRUE` by default, we don't actually need to list them in the function, just
those we want to switch to `FALSE` or those that require additional arguments.

```{r cleanup data}
motive_cleaned <-
  motive_data %>%
  clean_viewr(
    desired_percent = 70,
    rename_viewr_characters = FALSE,
    separate_trajectories = FALSE,
    get_full_trajectories = FALSE
  )

## Quick plot
plot(motive_cleaned$position_length,
     motive_cleaned$position_width,
     asp = 1)
```

### Inspecting the data We've cleaned up our data and are ready to link
individual data points into continuous trajectories. Deciding exactly how to
define a trajectory will depend largely on the quality of data collected and the
resolution required to answer a given question about the data. If the data is
fairly continuous and high resolution is required, the default 
`max_frame_gap = 1` will not allow any frame gaps--movement must be continuous.

```{r max_frame_gap_1}
motive_mfg1 <- 
  motive_cleaned %>% 
  separate_trajectories(
    max_frame_gap = 1
  )

## Quick plot
plot(motive_mfg1$position_length,
     motive_mfg1$position_width,
     asp = 1, col = as.factor(motive_mfg1$file_sub_traj))

## How many trajectories do we end up with?
length(unique(motive_mfg1$file_sub_traj))
```

Plotting each trajectory reveals that perhaps some trajectories that should be
continuous have been split into two or more separate trajectories because of a
dropped frame or two. This data may require relaxing the stringent no-frame-gaps
requirement in order to link data points that should go together into a single
trajectory. Let's try setting `max_frame_gap = 5`.

```{r max_frame_gap_5}
motive_mfg5 <- 
  motive_cleaned %>% 
  separate_trajectories(
    max_frame_gap = 5
  )

## Quick plot
plot(motive_mfg5$position_length,
     motive_mfg5$position_width,
     asp = 1, col = as.factor(motive_mfg5$file_sub_traj))

## How many trajectories do we end up with?
length(unique(motive_mfg5$file_sub_traj))
```

By increasing the allowable frame gap, more chunks of data have been linked into
trajectories and so our trajectory count has dropped from 335 to 224. Because it
is hard to see all the trajectories piled up on top of each other, let's go
ahead and run this through `get_full_trajectories()` to clean out bits of data
that do not span our region of interest. To inspect each trajectory
individually, we can then use the `plot_viewr_trajectories()` function to take a
closer look at the quality of each trajectory. This can be computationally
expensive depending on your data set.

```{r get_full_trajectories}
motive_mfg5_full <- 
  motive_mfg5 %>% 
  get_full_trajectories(
    span = .6
  )

## Quick plot
plot(motive_mfg5_full$position_length,
     motive_mfg5_full$position_width,
     asp = 1, col = as.factor(motive_mfg5_full$file_sub_traj))

## How many trajectories do we end up with?
length(unique(motive_mfg5_full$file_sub_traj))

## Plot each trajectory
plot_viewr_trajectories(motive_mfg5_full,
                        plot_axes = c("length", "width"),
                        multi_plot = TRUE)
```

### Visualize frame gap choice `pathviewR` has several tools to help users
decide what frame gap allowances may be appropriate depending on their data and
resolution needs. `visualize_frame_gap_choice()` runs the
`separate_trajectories()` function over the same data set as many times as the
user would like via the `loop` argument, each time with a different
`max_frame_gap` allowance.

```{r visualize_frame_gap_choice}
motive_cleaned %>% 
  visualize_frame_gap_choice(
    loops = 20
  )
```

The output of `visualize_frame_gap_choice()` is a tibble and plot of the number
of trajectories after running `separate_trajectories()` with `max_frame_gap =
1`, `max_frame_gap = 2`, `max_frame_gap = 3`, etc. We can see that as the frame
gap allowance increases, more bits of data are being linked into continuous
trajectories and thus the total number of trajectories decreases. The vertical
line on the plot indicates the "elbow" of the plot or the point at which counts
of trajectories appear to stabilize and increases in the `max_frame_gap`
allowance no longer effect the trajectory count very much.

### Autodetect
Setting `max_frame_gap = "autodetect"` rather than a numeric value uses the
"elbow" of the plots from `visualize_frame_gap_choice()` to guesstimate the best
value(s) for `max_frame_gap`. In addition to automatically selecting the
`max_frame_gap` depending on the data, "autodetect" does so on a per subject
basis rather than applying the same allowable frame gap to all data in the data
set. The cap on how high the `max_frame_gap` can go is defined as a proportion
of the capture frame rate and set by the `frame_rate_proportion` argument, which
defaults to `.10`.

```{r max_frame_gap_auto}
motive_auto <- 
  motive_cleaned %>% 
  separate_trajectories(
    max_frame_gap = "autodetect",
    frame_rate_proportion = 0.1,
    frame_gap_messaging = TRUE,
    frame_gap_plotting = TRUE
  )

## How many trajectories do we end up with?
length(unique(motive_auto$file_sub_traj))
```

Our sample data has 3 subjects so with `frame_gap_messaging = TRUE`, the
`max_frame_gap` for each subject will be reported (the default is `FALSE`).
`frame_gap_plotting = TRUE` will display the elbow plots for each subject, but
also defaults to `FALSE`.
