---
title: "Estimating visual perceptions from tracking data"
author: "Eric R. Press" 
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating visual perceptions from tracking data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{magrittr}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`pathviewR` includes functions that estimate visual perceptions based on the 
distance between the subject/observer and visual stimuli on the walls of the 
experimental tunnel.

## Loading packages
We'll need to start by loading `pathviewR` as well as a couple `tidyverse` 
packages for visualizing the data.

```{r package_loading, message=FALSE, warning=FALSE}
library(pathviewR)
library(ggplot2)
library(magrittr)
```


## Data preparation
Data objects must be prepared as desribed in [the Data Import and Cleaning vignette](https://vbaliga.github.io/pathviewR/articles/data-import-cleaning.html) 
pipeline prior to their use in these functions. For a detailed description of 
these functions, please see the linked vignette. 

Let's work with a few example datasets included in `pathviewR`. 
`pathviewR_motive_example_data.csv` is a .csv file exported from `Motive`. 
`pathviewR_flydra_example_data.mat` is a .mat file exported from `Flydra`.
`pathviewR` contains an all-in-one cleaning function `clean_viewr` for more 
coarse-grained data cleaning tasks. We will use this function in the examples 
below. 


```{r}
## motive data set
motive_data <- # import
  read_motive_csv(
    system.file("extdata", "pathviewR_motive_example_data.csv",
                package = 'pathviewR')
  )

motive_full <-
  motive_data %>%
  clean_viewr(
    relabel_viewr_axes = TRUE,
    gather_tunnel_data = TRUE,
    trim_tunnel_outliers = TRUE,
    standardization_option = "rotate_tunnel",
    select_x_percent = TRUE,
    desired_percent = 50,
    rename_viewr_characters = FALSE,
    separate_trajectories = TRUE,
    max_frame_gap = "autodetect",
    get_full_trajectories = TRUE,
    span = 0.95
  )
```
```{r}
## flydra data set
flydra_data <- # import
  read_flydra_mat(
    system.file("extdata", "pathviewR_flydra_example_data.mat",
                package = 'pathviewR'),
    subject_name = "birdie_wooster")

flydra_full <- # clean
  flydra_data %>%
  clean_viewr(
    relabel_viewr_axes = FALSE,
    gather_tunnel_data = FALSE,
    trim_tunnel_outliers = FALSE,
    standardization_option = "redefine_tunnel_center",
    length_method = "middle",
    height_method = "user-defined",
    height_zero = 1.44,
    get_velocity = FALSE,
    select_x_percent = TRUE,
    rename_viewr_characters = FALSE,
    separate_trajectories = TRUE,
    get_full_trajectories = TRUE
  )
```

### Add experiment information with `insert_treatments()`
Now that our objects have been cleaned, we will use `insert_treatments()` to add
information about the experiments that are necessary for calculating visual 
perceptions.  

#### Motive example
The data within `motive_full` were collected from birds flying through a
V-shaped tunnel in which the origin `(0,0,0)` was set to the height of the
perches that sit 0.3855m above the height of the vertex. The lateral walls were
angled ±45˚ from the vertical axis. The visual stimulus on the positive side of
the tunnel (where `position_width` values > 0) were horizontal sine wave
gratings with a cycle length of 0.1m on the screen. The visual stimulus on the
negative side of the tunnel (where `position_width` values < 0) were vertical
sine wave gratings with a cycle length of 0.2m on the screen.

Therefore we will use the following code:
```{r}
motive_V <- 
  motive_full %>% 
  insert_treatments(vertex_height = -0.3855,
                    vertex_angle = 45,
                    stim_param_pos = 0.1,
                    stim_param_neg = 0.2,
                    treatment = "latB")
```
Our object now has the variables, `vertex_height`, `vertex_angle`, 
`stim_param_pos`, and `stim_param_neg` which are needed to calculate visual 
perceptions. The variable `treatment` has also been included and this 
information has been stored in the object's metadata.
```{r}
motive_V
```

#### Flydra example
The data within `flydra_full` were collected from birds flying in a rectangular
tunnel i.e. a box where the positive, negative, and front walls were 0.5m from
the origin. The visual stimuli were the same as in the motive example, though
now with a stimulus on the front wall of 0.2m cycle length.

```{r}
flydra_box <- # prep for calculations
  flydra_full %>%
  insert_treatments(pos_wall = 0.5,
                    neg_wall = 0.5,
                    front_wall = 0.5,
                    stim_param_pos = 0.1,
                    stim_param_neg = 0.2,
                    stim_param_front = 0.2,
                    treatment = "latB")
```


## Calculating spatial frequency
To calculate the spatial frequency of the visual stimuli as perceived by the
subject some distance from the stimuli, we will use `calc_sf_V()` and
`calc_sf_box` depending on the configuration of the tunnel for each experiment. 

```{r}
motive_V_sf <- 
  motive_V %>%
  calc_sf_V(simplify_output = FALSE)

## The resulting object contains 10 new variables which are values involved in 
## the calculation of spatial frequency.
motive_V_sf
```
```{r}
flydra_box_sf <- 
  flydra_box %>% 
  calc_sf_box()
flydra_box_sf
```
`simplify_output = TRUE` returns an object containing the 4 new variables
`min_dist_pos`, `mind_dist_neg`, `sf_pos`, `sf_neg`. Note for `calc_sf_box`,
there is no need for a simplify_output argument.

## Calculating visual angles
To calculate an estimation of the visual angles perceived by the subject, we
will use `calc_vis_angle_V` and not `calc_vis_angle_box` because the data was
collected in a V-shaped tunnel.

```{r}
motive_V_angle <- 
  motive_V %>% 
  calc_vis_angle_V(simplify_output=FALSE)

## The resulting object contains 12 new variables which are values involved in the calculation of visual angles. 
motive_V_angle 
```
```{r}
flydra_box_angle <- 
  flydra_box %>% 
  calc_vis_angle_box()
```
`simplify_output=TRUE` returns an object containing the 4 new variables
`min_dist_pos`, `min_dist_neg`, `vis_angle_pos_deg`, and `vis_angle_neg_deg`.


## Visualizing the calculations
Visualizing the calculations provides an more intuitive understanding of how 
these visual perceptions change as the subject moves throughout the tunnel.
**Note:** The axes of the following examples differ between the motive and
flydra data sets. These plots were generated to best visualize the perceptual
calculations

### Spatial frequency
**Motive**
```{r motive_V_sf_pos, fig.height=4, fig.width=7}
ggplot(motive_V_sf, aes(x = position_width, y = position_height)) +
  geom_point(aes(color = sf_pos), shape=1, size=3) +
  coord_fixed() +
  geom_segment(aes(x = 0,         # dimensions of the positive wall
                  y = -0.3855,
                  xend = 0.5869,
                  yend = 0.2014)) +
  geom_segment(aes(x = 0,         # dimensions of the negative wall
                   y = -0.3855,
                   xend = -0.5869,
                   yend = 0.2014))

```
We can see that as the position of the subject is closer to the right (positive) wall, the perception of that stimulus is with smaller spatial frequency. 
```{r motive_V_sf_neg, fig.height=4, fig.width=7}
ggplot(motive_V_sf, aes(x = position_width, y = position_height)) +
  geom_point(aes(color = sf_neg), shape=1, size=3) +
  coord_fixed() +
  geom_segment(aes(x = 0,         # dimensions of the positive wall
                  y = -0.3855,
                  xend = 0.5869,
                  yend = 0.2014)) +
  geom_segment(aes(x = 0,         # dimensions of the negative wall
                   y = -0.3855,
                   xend = -0.5869,
                   yend = 0.2014))

```
A different pattern is found by visualizing the spatial frequencies observed
from the left (negative) wall. 


**Flydra**
```{r flydra_box_sf_pos, fig.height=4, fig.width=7}
ggplot(flydra_box_sf, aes(x = position_width, y = position_length)) +
  geom_point(aes(colour = sf_pos), size = 2) + 
  coord_fixed() +
  geom_segment(aes(x = -0.5,        # negative wall
                   y = -0.5,
                   xend = -0.5,
                   yend = 0.5)) +
  geom_segment(aes(x = 0.5,         # positive wall
                   y = -0.5,
                   xend = 0.5,
                   yend = 0.5))
```
```{r flydra_box_sf_neg, fig.height=4, fig.width=7}
ggplot(flydra_box_sf, aes(x = position_width, y = position_length)) +
  geom_point(aes(colour = sf_neg), size = 2) + 
  coord_fixed() +
  geom_segment(aes(x = -0.5,        # negative wall
                   y = -0.5,
                   xend = -0.5,
                   yend = 0.5)) +
  geom_segment(aes(x = 0.5,         # positive wall
                   y = -0.5,
                   xend = 0.5,
                   yend = 0.5))
```



### Visual angles
**Motive**
```{r motive_V_angle_pos, fig.height=4, fig.width=7}
ggplot(motive_V_angle, aes(x = position_width, y = position_height)) +
  geom_point(aes(color = vis_angle_pos_deg), shape=1, size=3) +
  coord_fixed() +
  geom_segment(aes(x = 0,         # dimensions of the positive wall
                  y = -0.3855,
                  xend = 0.5869,
                  yend = 0.2014)) +
  geom_segment(aes(x = 0,         # dimensions of the negative wall
                   y = -0.3855,
                   xend = -0.5869,
                   yend = 0.2014))
```

By displaying the visual angles (in degrees) created by the stimulus on the
positive wall, we can see that the subject perceives larger visual angles the
closer it gets to the positive wall.


Now displaying visual angles perceived on the negative wall. 
```{r motive_V_angle_neg, fig.height=4, fig.width=7}
ggplot(motive_V_angle, aes(x = position_width, y = position_height)) +
  geom_point(aes(color = vis_angle_neg_deg), shape=1, size=3) +
  coord_fixed() +
  geom_segment(aes(x = 0,         # dimensions of the positive wall
                  y = -0.3855,
                  xend = 0.5869,
                  yend = 0.2014)) +
  geom_segment(aes(x = 0,         # dimensions of the negative wall
                   y = -0.3855,
                   xend = -0.5869,
                   yend = 0.2014))
```
This shows that the subject perceives larger visual angles the closer it gets to
the negative wall. 

**Flydra**
```{r flydra_box_angle_pos, fig.height=4, fig.width=7}
ggplot(flydra_box_angle, aes(x = position_width, y = position_length)) +
  geom_point(aes(colour = vis_angle_pos_deg), size = 2) + 
  coord_fixed() +
  geom_segment(aes(x = -0.5,        # negative wall
                   y = -0.5,
                   xend = -0.5,
                   yend = 0.5)) +
  geom_segment(aes(x = 0.5,         # positive wall
                   y = -0.5,
                   xend = 0.5,
                   yend = 0.5))
```
```{r flydra_box_angle_neg, fig.height=4, fig.width=7}
ggplot(flydra_box_angle, aes(x = position_width, y = position_length)) +
  geom_point(aes(colour = vis_angle_neg_deg), size = 2) + 
  coord_fixed() +
  geom_segment(aes(x = -0.5,        # negative wall
                   y = -0.5,
                   xend = -0.5,
                   yend = 0.5)) +
  geom_segment(aes(x = 0.5,         # positive wall
                   y = -0.5,
                   xend = 0.5,
                   yend = 0.5))
```



## Future steps...
### Orientation 
### Optic flow estimation
